---
title: "Exploring a Clinical Trial Dataset"
subtitle: "Session 1: Summary Statistics and Visualisation"
format:
  html:
    code-annotations: hover
    code-fold: true
    code-fold-label: "Show Code"
    toc: true
    toc-depth: 3
    css: styles.css
editor: visual
---

```{r setup, include=FALSE}

library(tidyverse)

data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/main/datasets/sbp_demo.csv",
  show_col_types = FALSE
)

bp_data <- data |>
  filter(group %in% c("Placebo", "High Dose"))
bp_data

```

## Background

High blood pressure is a major risk factor for cardiovascular disease. Researchers have developed a new medication called **Hypotensol** and want to determine whether it reduces blood pressure more effectively than a placebo.

To investigate this, they conducted a randomised clinical trial.

::: highlight
### Research question

> Do patients receiving the **High Dose Hypotensol** treatment have lower post-treatment systolic blood pressure than patients receiving the **Placebo** treatment?
:::

::: callout-important
## Think like a researcher

We want to compare the **High Dose** and **Placebo** treatment groups.

Each patient has a post-treatment blood pressure measurement.

**What information would help us compare these two groups?**
:::

To compare the two treatment groups, it is useful to know:

- how many patients are in each group;
- the average post-treatment blood pressure in each group; and
- whether the blood pressure values are fairly similar or vary a lot within each group.

These are called **summary statistics**.

In this activity, we will calculate:

- **`n`** = the number of patients in each group;
- the **mean** = the average post-treatment systolic blood pressure; and
- the **standard deviation (SD)** = how much the blood pressure values vary within each group.

## (1) Calculate Summary Statistics for Each Group

```{r}
group_summary <- bp_data |>
  group_by(group) |>
  summarise(
    n = n(),
    Mean_SBP = mean(sbp_post),
    SD_SBP = sd(sbp_post)
  )

group_summary
```

- The `group_by()` function tells R to split the dataset into groups based on the variable `group`.

  - In our dataset, this creates two groups:

    - High Dose

    - Placebo

- The `summarise()` function calculates summary statistics for each group.

  - Here we calculate:
    - `n()` = number of patients in each treatment group
    - `mean()` = average post-treatment blood pressure
    - `sd()` = standard deviation, which measures how spread out the blood pressure values are

::: {.callout-note collapse="true"}
## Optional: Rounding the output

By default, R displays several decimal places.

To make the summary table easier to read, we can use the `round()` function.

For example:

```{r}
group_summary <- bp_data |>
  group_by(group) |>
  summarise(
    n = n(),
    Mean_SBP = round(mean(sbp_post), 1),
    SD_SBP = round(sd(sbp_post), 1)
  )

group_summary
```

The `round(x, 1)` function rounds a number to **one decimal place**.

For example:

- `125.12345` becomes `125.1`
- `8.6789` becomes `8.7`
:::

**Questions**

- Which group has the lower mean blood pressure?

- How many patients are in each group?

- Which group shows more variability in blood pressure values?

::: {.callout-tip collapse="true"}
## Interpreting the Summary Table

The summary table contains three useful pieces of information.

- **`n`** tells us how many patients are in each treatment group.

- **`Mean_SBP`** tells us the average post-treatment systolic blood pressure for each group.

- **`SD_SBP`** (standard deviation) tells us how much the blood pressure values vary within each group. A smaller SD indicates that patients have more similar blood pressure values, whereas a larger SD indicates greater variation.
:::

::: result-box
### Key finding

The **High Dose** group has a lower average post-treatment systolic blood pressure than the **Placebo** group.

This finding is consistent with our research hypothesis.
:::

## Optional: Checking for Missing Values

::: {.callout-note collapse="true"}
## Note: What if some values are missing? - How to check and what to do

Before calculating statistics, researchers should check whether a variable contains missing values. Missing information can affect the results of an analysis.

In R, missing values are shown as `NA`.

To check whether `sbp_post` contains missing values, we can use:

```{r}
sum(is.na(data$sbp_post))
```

If the result is `0`, there are no missing values in this variable.

If a variable does contain missing values, many R functions allow us to ignore them using `na.rm = TRUE`.

For example:

```{r}
mean(data$sbp_post, na.rm = TRUE)
```

This tells R:

> Calculate the mean after removing missing values.
:::

## (2) Visualising the Results

> **Why create a graph?**
>
> - The summary table allows us to compare the treatment groups numerically.
>
> - A graph allows us to compare the treatment groups more quickly and makes differences easier to see.
>
> - Researchers often create graphs because they make differences between groups easier to see.

Let's create a simple **bar chart** showing the **mean post-treatment systolic blood pressure** for each treatment group.

```{r}
ggplot(group_summary,
       aes(x = group,
           y = Mean_SBP)) +
  geom_col()
```

### Understanding the code

- `ggplot()` starts a new graph using the dataset `group_summary`.

  - `aes()` (short for *aesthetics*) tells R which variables to display:

    - `x = group` places the treatment groups on the x-axis.

    - `y = Mean_SBP` places the mean blood pressure on the y-axis.

- `geom_col()`= creates a bar chart using values that have already been calculated.

  - In our case, the heights of the bars come from the variable: `Mean_SBP`

::: {.callout-note collapse="true"}
## The `+` symbol in ggplot

Unlike many R functions, graphs are built one layer at a time.

The `+` symbol tells R to **add another layer** to the graph.

For example:

- `geom_col()` adds the bars.
- `geom_text()` adds the labels.
- `labs()` adds the axis titles.

You can think of `+` as meaning **"and then add..."**.
:::

## Improving the graph

The graph is informative, but we can make it easier to interpret by:

1.  Adding colours;

2.  Displaying the mean value above each bar; and

3.  Labelling the axes.

```{r}
ggplot(group_summary,
       aes(x = group,
           y = Mean_SBP)) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(label = round(Mean_SBP, 1)),
    vjust = -0.5
  ) +
  labs(
    x = "Treatment group",
    y = "Mean post-treatment systolic blood pressure (mmHg)"
  )
```

::: {.callout-note collapse="true"}
## Optional: 1. Changing the bar colour

The argument `fill = "steelblue"` changes the colour of the bars.

You can replace `"steelblue"` with many other colour names, for example:

- `"tomato"`
- `"darkgreen"`
- `"gold"`
- `"purple"`

Changing the colour does **not** change the data; it simply improves the appearance of the graph.
:::

::: {.callout-note collapse="true"}
## Optional: 2. Displaying the mean value above each bar

The function `geom_text()` adds text to a graph.

Here, it displays the mean blood pressure value above each bar.

- `label = round(Mean_SBP, 1)` displays the mean rounded to one decimal place.
- `vjust = -0.5` positions the label slightly above the bar.
:::

::: {.callout-note collapse="true"}
## Optional: 3. Labelling the axes

- By default, ggplot uses the variable names as axis labels.

- The `labs()` function allows us to replace these with clearer, more descriptive labels.

In this example:

``` r
labs(
  x = "Treatment group",
  y = "Mean post-treatment systolic blood pressure (mmHg)"
)
```

- `x =` sets the label for the **x-axis**.

- `y =` sets the label for the **y-axis**.

Giving informative axis labels makes graphs easier to understand and interpret.
:::

### Interpreting the Plot

Look at the graph.

**Questions**

- Which treatment group has the lower mean post-treatment blood pressure?
- Approximately how much lower is it than the other group?
- Does the graph support the research hypothesis?

::: {.result-box collapse="true"}
### Key finding

Both the summary table and the bar chart suggest that patients receiving the **High Dose** treatment had lower post-treatment systolic blood pressure than patients receiving the **Placebo** treatment.

This finding is consistent with our research hypothesis.
:::

### But...

Although the High Dose group appears to have a lower average blood pressure, we do not yet know whether the observed difference is **statistically significant**.

**Could this difference have occurred simply by chance?**

To answer this question, we need to perform a **hypothesis test**.

::: callout-note
## Looking ahead to Session 2

Today, we used **summary statistics** and **data visualisation** to explore our research question.

In **Session 2**, we will learn how to use **hypothesis tests** to determine whether the difference we observed between the High Dose and Placebo groups is statistically significant.

We will also learn how to choose the appropriate statistical test based on the research question and the types of variables being analysed.
:::
