---
title: "Exploring a Clinical Trial Dataset"
subtitle: "Session 1: Summary Statistics and Visualisation"
format:
  html:
    code-annotations: hover
    code-fold: false
    toc: true
    toc-depth: 3
    css: styles.css

execute:
  echo: true
  warning: false
  message: false

editor: visual
---

```{r setup, include=FALSE}

library(tidyverse)
```

# Exploring a Clinical Trial Dataset

In this activity, you will use R to explore data from a fictional clinical trial investigating a new blood pressure medication called **Hypotensol**. You will answer a research question by identifying the relevant variables, summarising the data, and creating simple visualisations.

## Background

High blood pressure is a major risk factor for cardiovascular disease. Researchers have developed a new medication called **Hypotensol** and want to determine whether it reduces blood pressure more effectively than a placebo.

To investigate this, they conducted a randomised clinical trial.


### Research question

> Do patients receiving the **High Dose Hypotensol** treatment have lower post-treatment systolic blood pressure than patients receiving the **Placebo** treatment?


## Dataset

To answer this question, we will analyse data from a fictional clinical trial involving **90 patients** with high blood pressure.

Participants were randomly assigned to one of three treatment groups:

1.  Placebo
2.  Low Dose Hypotensol (25 mg)
3.  High Dose Hypotensol (50 mg)

Patients were recruited from:

- NHS Lothian

- NHS Greater Glasgow

### Variables in the Dataset

| Variable Name | Description |
|-------------------------------|-----------------------------------------|
| `id` | Patient ID number (anonymised unique identifier) |
| `group` | Treatment group (Placebo / Low Dose / High Dose) |
| `sbp_pre` | Systolic blood pressure before treatment (mmHg) |
| `sbp_post` | Systolic blood pressure after treatment (mmHg) |
| `improved` | Did the patient improve? (Yes / No) |
| `age` | Age of patient in years |
| `sex` | Sex of patient (Male / Female) |
| `weight_kg` | Weight at baseline in kilograms |
| `bmi` | Body Mass Index (kg/m²) |
| `drug` | Drug administered (Hypotensol / None) |
| `dose_mg` | Dose of drug administered in milligrams |
| `post_check_weeks` | Weeks after treatment when post-treatment BP was measured |
| `home_sbp_avg` | Average home blood pressure reading after treatment (mmHg) |
| `followup_1y` | 1-year follow-up status (Completed / Missed) |
| `clinic` | NHS clinic where the patient was treated |

::: definition-box
### Researcher's mindset

Every statistical analysis begins with a research question. The research question determines:

- which variables we need;

- whether those variables are categorical or numerical; and

- which statistical methods are appropriate.
:::


### Learning objectives

By the end of this activity, you should be able to:

- import a dataset into R;

- inspect the structure of a dataset;

- identify categorical and numerical variables;

- use R to summarise and visualise data;

- interpret descriptive statistics in the context of a research question.


::: example-box
### Identifying the variables

Before opening R, let's return to our research question.

> Do patients receiving the **High Dose Hypotensol** treatment have lower post-treatment systolic blood pressure than patients receiving the **Placebo** treatment?

Read the question carefully.

**Which variables from the dataset do we need to answer it?**

> Ask yourself:
>
> - Which variable tells us **who received each treatment**?
> - Which variable contains the **outcome we want to compare**?
:::

::: {.callout-tip collapse="true"}
## Solution

From the table above, we can identify two key variables.

| Variable | Why do we need it? | Type |
|--------------------|----------------------------------|------------------|
| `group` | Identifies which treatment each patient received (Placebo, Low Dose or High Dose). | Categorical |
| `sbp_post` | Records each patient's post-treatment systolic blood pressure. | Numerical |

Notice that we do **not** need every variable in the dataset.
:::

# Working in RStudio

Now that we know the research question and have identified the variables we need, it's time to begin the analysis in **RStudio**.

Let's open [**Getting Started with RStudio**](Getting_started_with_RStudio.html) guide before continuing.

Throughout this activity, you will work in a downloadable **R Markdown (`.Rmd`) template**. The template contains the code and exercises we will complete together during the session.

::: checklist-box
### Getting Started

Before we begin:

1.  📥 **Download the [Day 1 Student Template (.Rmd)](student-templates/Session1_student_template_Clinical_Trial.Rmd).**
2.  Save it in a folder such as **Documents → Summer School 2026**.
3.  Open the file in **RStudio**.
4.  Keep both this webpage and RStudio open during the session.
:::

::: definition-box
### How will we use the template?

The template is your working document for today's session.

As we move through the activity, you will:

- run the code provided in each section;
- observe and interpret the output produced by R;
- answer short questions about the results; and
- complete a few "Your Turn" exercises by writing your own code.

By the end of the session, you will have a completed analysis that you can save and use as a reference.
:::

## Step 1: Import the Dataset

> Before we can analyse the data, we first need to import it into R.

The first step is to import the dataset into R.

The dataset is stored online as a **CSV** file, so R can read it directly from the web.

```{r}
library(tidyverse)

data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/main/datasets/sbp_demo.csv",
  show_col_types = FALSE
)
```

## Step 2: Did R import the data correctly?

```{r}
head(data)
```

Questions:

- What is the number of rows and columns?

- What does one row represent?

- What does one column represent?

Then look at :

- `<chr>` = text (categorical variable)

- `<dbl>` = number (numerical variable)

## Step 3: What variables might answer our research question?

### Research question:

> Do patients receiving the High Dose Hypotensol treatment have lower post-treatment systolic blood pressure than patients receiving the Placebo treatment?

Which variable tells us:

- treatment group?

- post-treatment blood pressure?

```{r echo=FALSE}

library(DT)

DT::datatable(
  data,
  class = "compact stripe hover",
  options = list(
    pageLength = 8,
    scrollX = TRUE,
    autoWidth = FALSE
  )
)

```

::: {.callout-tip collapse="true"}
## Solution

From the table above, we can identify two key variables.

| Variable | Why do we need it? | Type |
|--------------------|----------------------------------|------------------|
| `group` | Identifies which treatment each patient received (Placebo, Low Dose or High Dose). | Categorical |
| `sbp_post` | Records each patient's post-treatment systolic blood pressure. | Numerical |

Notice that we do **not** need every variable in the dataset.
:::

## Step 4: How many patients are in each treatment group?

### Count how many patients are in each treatment group.

The function`count()`

- counts how many observations belong to each category of the variable `group`.

- mainly used with **categorical variables**.

In our dataset, the variable `group` contains the treatment assigned to each patient:

- Placebo

- Low Dose

- High Dose

#### The syntax

```         
count(dataset, variable)
```

where:

- `dataset` = the dataset we want to analyse;

- `variable` = the categorical variable we want to count.

#### In our example

```{r}
count(data, group)
```

- `count()` = count the number of observations

- `data` = the dataset we want to analyse

- `group` = the variable we want to count

- `n` = shows the number of patients in each group

#### Interpreting the output: Why is this useful?

The output shows the number of patients in each treatment group.

Before analysing data, researchers often ask

- How many patients are in each treatment group?
- Are the treatment groups balanced?
- Are any treatment groups missing?

In this clinical trial, each treatment group contains \*\*30 patients\*\*, so the groups are perfectly balanced.

### Your turn! Explore another categorical variable

Use the `count()` function to find out how many males and females are in the dataset.

```{r}
# Write your code here
```

**Questions**\
\
- How many males are in the dataset?\
\
- How many females are in the dataset?\
\
- Are there roughly equal numbers of males and females in the dataset?

::: {.callout-tip collapse="true"}
## Solution

```{r}
count(data, sex)
```
:::

## Step 5: Create a new dataset (subset) for our research question

Let's return to our research question:

> Do patients receiving the **High Dose Hypotensol** treatment have lower post-treatment systolic blood pressure than patients receiving the **Placebo** treatment?

Our dataset contains **three** treatment groups:

- Placebo

- Low Dose

- High Dose

However, our research question only compares **High Dose** and **Placebo**.

Therefore, before analysing the data, we will create a **new dataset** containing only the patients relevant to our research question.

### The `filter()` function

The `filter()` function selects only the rows that satisfy a condition.

#### The syntax

```         
filter(variable == value)
```

or, when keeping multiple values,

```         
filter(variable %in% c(value1, value2))
```

#### In our example

**The filter() function allows us to select observations that meet specific conditions.**

```{r}
bp_data <- data |>
  filter(group %in% c("Placebo", "High Dose"))
bp_data
```

We can read this code as: Take the dataset `data`, keep only the patients in the Placebo and High Dose groups, and store the result in a new dataset called `bp_data`.

### Reading the code

Let's read the code from left to right.

- `data` = start with the original dataset.

- `|>` = pass the dataset to the next function.

- `filter()` = keep only the rows that satisfy our condition.

- `group %in% c("Placebo", "High Dose")` = keep only patients in the **Placebo** or **High Dose** treatment groups.

- `bp_data <-` = store the filtered dataset in a new object called `bp_data`.

::: column-margin
#### The Assignment Operator: `<-`

- The symbol `<-` means: Store the result on the right-hand side in the object on the left-hand side.

- In our analysis : `bp_data <- data` means " Create a new object called `bp_data` and store the dataset `data` inside it."
:::

::: column-margin
#### The Pipe Operator: `|>`

- The symbol `|>` is called the **pipe operator**.

- it means: Take the result from the left and pass it to the function on the right.

```{r}
 data |>
    count(group)
```

can be read as:

> Take the dataset `data` and count the number of observations in each treatment group.
:::

### Why is this useful?

- Researchers rarely analyse every observation in a dataset.

- Instead, they select the subset of data that answers their research question.

- In this example, the **Low Dose** group is not relevant because our research question compares only the **High Dose** and **Placebo** groups.

## Step 6: Comparing Treatment Groups

Let's return to our research question:

> Do patients receiving the **High Dose Hypotensol** treatment have lower post-treatment systolic blood pressure than patients receiving the **Placebo** treatment?

To answer this question, we are comparing:

- a **categorical variable** (`group`), which identifies the treatment each patient received; and

- a **numerical variable** (`sbp_post`), which records each patient's post-treatment systolic blood pressure.

In the previous step, we created a new dataset called `bp_data`, which contains only the **High Dose** and **Placebo** groups. These are the only patients relevant to our research question.

**What should we calculate now?**

::: callout-important
## Think like a researcher

We want to compare the **High Dose** and **Placebo** groups.

Each patient has a post-treatment blood pressure value.

**What information would help you compare these two groups?**

Take a moment to think before looking at the solution.
:::

::: {.callout-tip collapse="true"}
## Solution

To compare the two groups, it is useful to know:

- how many patients are in each group;
- the average post-treatment blood pressure; and
- whether blood pressure values vary a lot within each group.

We can calculate these using summary statistics.
:::

------------------------------------------------------------------------

::: callout-tip
## Continue to the next activity

You have now:

- identified the variables needed to answer the research question;
- explored the dataset;
- counted observations in categorical variables; and
- created a dataset containing only the treatment groups of interest.

**Next:**\
[**Summary Statistics and Visualisation**](Clinical_trial_summary.html)
:::
