---
title: "Exploring a Clinical Trial Dataset"
subtitle: "Session 2: Hypothesis testing"
format:
  html:
    code-annotations: hover
    code-fold: true
    code-fold-label: "Show Code"
    toc: true
    toc-depth: 3
    css: styles.css
editor: visual
---

```{r setup, include=FALSE}

library(tidyverse)

data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/main/datasets/sbp_demo.csv",
  show_col_types = FALSE
)

bp_data <- data |>
  filter(group %in% c("Placebo", "High Dose"))
bp_data
```

## 1. Where we left off

In Session 1 we found that the **High Dose** group had a lower average post-treatment blood pressure than the **Placebo** group.

However, one important question remains.

> Is this difference large enough that we can attribute it to the treatment, or could it simply have occurred by chance?

To answer this question, we need a statistical test.

## 2. Why do we need a hypothesis test?

Although the High Dose group had a lower average post-treatment blood pressure than the Placebo group, averages calculated from a sample are rarely exactly the same as the averages in the whole population.

A hypothesis test helps us decide whether the difference we observed is likely to reflect a real treatment effect or whether it could simply be due to random sampling variation.

## 3. Choosing the right statistical test

Researchers do not choose statistical tests at random.

The choice depends on the research question and the type of data being analysed.

::: callout-important
## Think like a researcher

Our research question is:

> Do patients receiving the High Dose treatment have lower post-treatment systolic blood pressure than patients receiving the Placebo treatment?

Ask yourself:

- What is our outcome variable?
- Is it numerical or categorical?
- How many groups are we comparing?
:::

::: {.callout-tip collapse="true"}
## Solution

| Variable | Description | Type | Role |
|----|----|----|----|
| `sbp_post` | SBP after treatment | Numeric | Dependent variable (DV) |
| `group` | Treatment group (Placebo / High Dose) | Categorical (2 levels) | Independent variable (IV) |

Therefore, we need a statistical test that compares the means of two groups.
:::

::: {.callout-note appearance="simple" icon="info" title="What is an independent variable?"}
It's the thing you're **grouping by or comparing across** — in this case, treatment group.
:::

::: {.callout-tip collapse="true"}
### Summary Table of different test statistics for hypothesis testing

| Research question | Variables | Test |
|----|----|----|
| Compare two group means | One numeric + one categorical (2 groups) | t-test |
| Compare three or more group means | One numeric + one categorical (3+ groups) | ANOVA |
| Compare two categorical variables | Two categorical | Chi-square |
| Relationship between two numeric variables | Two numeric | Correlation |
:::

### The independent samples t-test

The independent samples t-test compares the average value of a numerical variable between two independent groups.

In our study:

- numerical variable: `sbp_post`
- groups: High Dose and Placebo

This is exactly the type of research question that a t-test is designed to answer.

> Example: Is systolic blood pressure (SBP) different for patients who took **High Dose Hypotensol** vs those who took **Placebo**?

## 4. Writing the hypotheses

- Every hypothesis test begins with a **research question**.

- The hypotheses are derived directly from that research question.

::: callout-tip
- The research question tells us **what we want to investigate**.
- The hypotheses translate that question into statements that we can test statistically.
:::

Our research question is:

> Do patients receiving the **High Dose** treatment have lower post-treatment systolic blood pressure than patients receiving the **Placebo** treatment?

From this research question, we can formulate two hypotheses.

1.  The **null hypothesis** says there is no difference between the groups.

2.  The **alternative hypothesis** says there is a difference, or that the difference goes in the direction predicted by the research question.

::: example-box
### Hypotheses for our research question

**Null hypothesis (H₀):**\
The High Dose and Placebo groups have the same average post-treatment systolic blood pressure.

**Alternative hypothesis (H₁):**\
The High Dose group has lower average post-treatment systolic blood pressure than the Placebo group.
:::

::: {.callout-note collapse="true"}
## Why a one-sided t-test?

There are two common types of independent samples t-test.

### 1. One-sided t-test

- A **one-sided t-test** is used when the research question predicts the **direction** of the difference.

For example:

> Do patients receiving the **High Dose** treatment have **lower** post-treatment blood pressure than patients receiving the **Placebo** treatment?

Here, we are specifically predicting that the High Dose group will have **lower** blood pressure.

### 2. Two-sided t-test

- A **two-sided t-test** is used when we only want to know whether the groups are **different**, without predicting which group will have the higher or lower values.

For example:

> Is there a difference in post-treatment blood pressure between the High Dose and Placebo groups?

This question does not predict the direction of the difference.

### Which one are we using?

Because our research question predicts that the **High Dose group will have lower blood pressure than the Placebo group**, we use a **one-sided t-test**.
:::

## 5. Running the t-test in R

### Preparing the data

- R compares the groups in the order they are stored.

- To match our research hypothesis, we want R to compare:

> **High Dose \< Placebo**

The following code sets the groups in this order.

So we set **High Dose** as the first group and **Placebo** as the second group.

```{r}

bp_data <- bp_data |>
  mutate(
    group = factor(
      group,
      levels = c("High Dose", "Placebo")
    )
  )
```

### Running the test

> Test whether the **High Dose** group has a lower mean `sbp_post` than the **Placebo** group.

```{r}
t.test(
  sbp_post ~ group,
  data = bp_data,
  alternative = "less"
)
```

### Understanding the code

- `sbp_post ~ group` tells R to compare post-treatment blood pressure across the treatment groups.

- `data = bp_data` tells R to use the filtered dataset containing only High Dose and Placebo patients.

- `alternative = "less"` tells R that our hypothesis is directional: we expect the High Dose group to have lower blood pressure than the Placebo group.

::: {.callout-note collapse="true"}
## Optional: Understanding the `~` symbol

In R, the symbol `~` can be read as **"by"** or **"grouped by."**

For example:

```         
sbp_post ~ group
```

can be read as:

> Compare post-treatment systolic blood pressure **by** treatment group.
:::

## 6. Interpreting the t-test output

Running `t.test()` produces several pieces of information.

Before reading the explanation below, see if you can identify the following in the output:

- the **group means**

- the **t statistic**;

- the **degrees of freedom (df)**

- the **p-value**; and

- the **confidence interval**.

```{r}

t.test(sbp_post ~ group,
  data = bp_data,
  alternative = "less")

```

**Where can you find each value?**

| Output | What does it tell us? |
|----|----|
| `t` | The **t-statistic** |
| `df` | The **degrees of freedom** |
| `p-value` | Whether the difference is statistically significant |
| `95 percent confidence interval` | A range of plausible values for the true difference between the group means |
| `mean in group High Dose` | The average SBP for the High Dose group |
| `mean in group Placebo` | The average SBP for the Placebo group |

### Step 1: Compare the group means

The output reports the mean post-treatment blood pressure for each group.

- mean in group High Dose 125.12

- mean in group Placebo 143.13

**Question**

Which treatment group has the lower average blood pressure?

::: {.callout-tip collapse="true"}
## Answer

The **High Dose** group has the lower average post-treatment blood pressure.

This agrees with the descriptive statistics we calculated in Session 1.
:::

### Step 2: Look at the p-value

The p-value helps us decide whether the observed difference is likely to reflect a real treatment effect or whether it could simply be due to chance.

::: {.callout-note appearance="simple"}
### What is a p-value?

The **p-value** tells us how compatible our data are with the idea that there is **no real difference** between the treatment groups.

A **small p-value** suggests that the observed difference is unlikely to be explained by chance alone.
:::

The p-value from our analysis is:

> **3.767 × 10⁻¹⁰**

This is much smaller than **0.05**.

::: {.callout-tip appearance="simple"}
### Decision rule

- If **p \< 0.05**, we reject the null hypothesis.

- If **p ≥ 0.05**, we do not reject the null hypothesis.
:::

### Step 3: What about the other numbers?

- The t-test output also reports a **t-statistic**, **degrees of freedom (df)** and a **confidence interval**.

- These values are important because they are used to calculate and describe the statistical result.

- For today's workshop, the **most important value to interpret is the p-value**, since it tells us whether the observed difference is statistically significant.

::: {.callout-note collapse="true"}
### Optional: What do these values mean?

- **t-statistic** – measures how large the difference between the group means is relative to the variation in the data.

- **Degrees of freedom (df)** – a value used when calculating the t-test.

- **Confidence interval** – gives a range of plausible values for the true difference between the group means. Because we performed a **one-sided t-test**, R reports a one-sided confidence interval.
:::

## 7. Conclusion

The t-test produced a **very small p-value** (`3.77 × 10⁻¹⁰`), which is much smaller than 0.05.

Therefore, we reject the null hypothesis.

The results provide strong evidence that patients receiving the **High Dose Hypotensol** treatment had lower average post-treatment systolic blood pressure than patients receiving the **Placebo** treatment.

### Reporting results

Researchers usually summarise their findings in a single sentence.

For example:

> An independent samples t-test found that patients in the High Dose group ($M = 125.12$) had significantly lower post-treatment systolic blood pressure than those in the Placebo group ($M = 143.13$), $t(57.44) = −7.36, p < .001$

::: {.callout-note collapse="true"}
### Understanding the reporting format

Research papers often report:

- $M$ = mean (average)

- $SD$ = standard deviation

- $t(df)$ = the t-statistic and its degrees of freedom

- $p$ = the p-value

This provides a concise summary of both the descriptive statistics (the group averages) and the inferential statistics (the hypothesis test).
:::

## 8. Key takeaway

In Session 1, we used **descriptive statistics** to describe the data.

In Session 2, we used **inferential statistics** to decide whether the observed difference between the treatment groups was statistically significant.

Together, these steps form a typical workflow for analysing data in a clinical trial.
