---
title: "Exploring a Clinical Trial Dataset"
author: "UoE Summer School 2026 Team "
date: "2026"
---

# Importing the data set

```{r}
library(tidyverse)

data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/main/datasets/sbp_demo.csv",
  show_col_types = FALSE
)

```

We have imported the dataset, but how do we know R has read it correctly?

## Step 1: Look at the first few rows

```{r}

head(data)
```

- What does each row represent?

- What does each column represent?

## Step 2: Ask R about the dataset

```{r}

glimpse(data)
```

- What is the number of rows and columns?

- What do you think these represent?

Then look at :

- `<chr>` = text (categorical variable)

- `<dbl>` = number (numerical variable)

## Step 3: What variables might answer our research question?

### Research question:

> Do patients receiving the High Dose Hypotensol treatment have lower post-treatment systolic blood pressure than patients receiving the Placebo treatment?

Which variable tells us:

- treatment group?

- post-treatment blood pressure?

## Step 4: Check the categories

### Count how many patients are in each treatment group.

The function`count()`

- counts how many observations belong to each category of the variable `group`.

- mainly used with **categorical variables**.

In our dataset, the variable `group` contains the treatment assigned to each patient:

- Placebo

- Low Dose

- High Dose

```{r}
count(data, group)
```

- `count()` = count the number of observations

- `data` = the dataset we want to analyse

- `group` = the variable we want to count

- `n` = shows the number of patients in each group

### Why is this useful?

Before analysing data, researchers often want to know:

- How many observations are in each category?

- Are the groups balanced?

- Are any categories missing?

In this clinical trial, the output tells us that the researchers assigned the same number of patients to each treatment group.

### Your turn! Explore another categorical variable

Use the `count()` function to find out how many males and females are in the dataset.

```{r}
# Write your code here
```

**Questions**\
\
- How many males are in the dataset?\
\
- How many females are in the dataset?\
\
- Are there roughly equal numbers of males and females in the dataset?

## Step 5: Filter High Dose and Placebo

Now we return directly to the research question:

> Do patients receiving the High Dose Hypotensol treatment have lower post-treatment systolic blood pressure than patients receiving the Placebo treatment?

Our research question compares the **High Dose** and **Placebo** groups. Do we need all three treatment groups?

### Create a Subset for the Research Question

Before we create our subset, let's look at two symbols that appear frequently in R.

#### The Assignment Operator: `<-`

- The symbol `<-` means: Store the result on the right-hand side in the object on the left-hand side.

  - In our analysis : `bp_data <- data` means " Create a new object called `bp_data` and store the dataset `data` inside it."

#### The Pipe Operator: `|>`

- The symbol `|>` is called the **pipe operator**.

- it means: Take the result from the left and pass it to the function on the right.

```{r}
 data |>
    count(group)
```

can be read as:

> Take the dataset `data` and count the number of observations in each treatment group.

### Creating a Subset

The dataset contains three treatment groups:

- Placebo

- Low Dose

- High Dose

However, our research question only compares:

- Placebo

- High Dose

Therefore, we will create a new dataset containing only these two groups.

**The filter() function allows us to select observations that meet specific conditions.**

```{r}
bp_data <- data |>
  filter(group %in% c("Placebo", "High Dose"))

```

- `filter()` = selects rows from a dataset.

- `group %in% c("Placebo", "High Dose")` = Keep only patients whose treatment group is either "Placebo" or "High Dose".

- `bp_data`= name of new dataset

We can read this code as: Take the dataset `data`, keep only the patients in the Placebo and High Dose groups, and store the result in a new dataset called `bp_data`.

## Step 6: Comparing Treatment Groups

Now we return directly to the research question:

> Do patients receiving the High Dose Hypotensol treatment have lower post-treatment systolic blood pressure than patients receiving the Placebo treatment?

We are comparing:

- a **categorical variable** (`group`)

- a **numerical variable** (`sbp_post`)

To begin, we can calculate the average blood pressure for each treatment group.

### (1) Calculate Summary Statistics for Each Group

```{r}
group_summary <- bp_data |>
  group_by(group) |>
  summarise(
    n = n(),
    Mean_SBP = mean(sbp_post),
    SD_SBP = sd(sbp_post)
  )

group_summary
```

- The `group_by()` function tells R to split the dataset into groups based on the variable `group`.

  - In our dataset, this creates two groups:

    - High Dose

    - Placebo

- The `summarise()` function calculates summary statistics for each group.

  - Here we calculate:
    - `n()` = number of patients in each treatment group
    - `mean()` = average post-treatment blood pressure
    - `sd()` = standard deviation, which measures how spread out the blood pressure values are

**Questions**

- Which group has the lower mean blood pressure?

- How many patients are in each group?

- Which group shows more variability in blood pressure values?

### Understanding the Summary Statistics

The table provides three important pieces of information about each treatment group.

- `n` tells us how many patients are in each group.

- `Mean_SBP` tells us the average post-treatment blood pressure for each group.

- `SD_SBP` (standard deviation) tells us how much variation there is in blood pressure values within a group.

A small standard deviation means that most patients have blood pressure values close to the group average. A large standard deviation means that patients differ more from one another.

When comparing treatment groups, we are usually interested in:

- whether one group has a lower average blood pressure than the other; and

- whether the blood pressure values are similarly spread out within each group.

For example, if one group has a standard deviation of 8 mmHg and another has a standard deviation of 20 mmHg, the second group shows much greater variation in blood pressure values.

The summary statistics suggest that the High Dose group has a lower average post-treatment blood pressure than the Placebo group. However, we do not yet know whether this difference is large enough to be considered statistically significant.

### (2) Plotting group means

```{r}
ggplot(group_summary,
       aes(x = group,
           y = Mean_SBP)) +
  geom_col()
```

**How do we read the code?** Take the summary table, place the treatment groups on the x-axis, place the mean blood pressure values on the y-axis, and draw a bar for each group.

- `geom_col()`= creates a bar chart using values that have already been calculated.

  - In our case, the heights of the bars come from the variable: `Mean_SBP`

- `aes()` = tells R: Put the treatment groups on the x-axis. Put the mean blood pressure values on the y-axis.

### Adding Labels to the Bars

```{r}
ggplot(group_summary,
       aes(x = group,
           y = Mean_SBP)) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(label = round(Mean_SBP, 1)),
    vjust = -0.5
  ) +
  labs(
    x = "Treatment group",
    y = "Mean post-treatment systolic blood pressure (mmHg)"
  )
```

- `geom_text()` = adds text to a plot

Here, it displays t**he mean blood pressure value above each bar**.

- `label = round(Mean_SBP, 1)` = display the mean rounded to 1 decimal place.

- `vjust = -0.5` = move the label slightly above the bar.

### Interpreting the Plot

The mean post-treatment blood pressure is:

- **125.1 mmHg** for the High Dose group

- **143.1 mmHg** for the Placebo group

The High Dose group has a lower average blood pressure than the Placebo group.

This finding is consistent with our hypothesis that patients receiving the High Dose treatment have lower post-treatment blood pressure than patients receiving the Placebo treatment.

### But...

Although the groups appear different, we do not yet know whether this difference is statistically significant.

**Could this difference have occurred simply by chance?**

To answer this question, we need to perform a hypothesis test.

## Note: Missing Values ? - How to check and what to do

Before calculating statistics, researchers should check whether a variable contains missing values. Missing information can affect the results of an analysis.

- `data$variable_name` = Selects a single variable from the dataset.

```{r}
data$sbp_post
```

- `is.na()` = Checks whether each value is missing.

```{r}
is.na(data$sbp_post)
```

- `FALSE` = value is present

- `TRUE` = value is missing

To count how many values are missing in the variable `sbp_post`.

```{r}
sum(is.na(data$sbp_post))
```

there are no missing values.

If you do have missing values:

- You can add `na.rm = TRUE` to many R functions, so that we can ignore missing values automatically

> "Remove missing values before calculating the mean."

```{r}
mean(data$sbp_post, na.rm = TRUE)
```
