---
title: "Exploring a Clinical Trial Dataset"
author: "UoE Summer School 2026 Team "
date: "2026"
---
# Part 1: Exploring the Dataset

## Importing the data set

```{r}
library(tidyverse)

data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/main/datasets/sbp_demo.csv",
  show_col_types = FALSE
)

```

We have imported the dataset, but how do we know R has read it correctly?

## Step 1: Look at the first few rows

```{r}

head(data)
```

- What does each row represent?

- What does each column represent?

## Step 2: Ask R about the dataset

```{r}

glimpse(data)
```

- What is the number of rows and columns?

- What do you think these represent?

Then look at :

- `<chr>` = text (categorical variable)

- `<dbl>` = number (numerical variable)

## Step 3: What variables might answer our research question?

### Research question:

> Do patients receiving the High Dose Hypotensol treatment have lower post-treatment systolic blood pressure than patients receiving the Placebo treatment?

Which variable tells us:

- treatment group?

- post-treatment blood pressure?

## Step 4: Count treatment groups

### Count how many patients are in each treatment group.

The function`count()`

- counts how many observations belong to each category of the variable `group`.

- mainly used with **categorical variables**.

```{r}
count(data, group)
```

- `count()` = count the number of observations

- `data` = the dataset we want to analyse

- `group` = the variable we want to count

- `n` = shows the number of patients in each group

### Why is this useful?

Before analysing data, researchers often want to know:

- How many observations are in each category?

- Are the groups balanced?

- Are any categories missing?


### Your turn! Explore another categorical variable

Use the `count()` function to find out how many males and females are in the dataset.

```{r}
# Write your code here
```

**Questions**\
\
- How many males are in the dataset?\
\
- How many females are in the dataset?\
\
- Are there roughly equal numbers of males and females in the dataset?

## Step 5: Filter High Dose and Placebo

Now we return directly to the research question:

> Do patients receiving the High Dose Hypotensol treatment have lower post-treatment systolic blood pressure than patients receiving the Placebo treatment?

Our research question compares the **High Dose** and **Placebo** groups. Do we need all three treatment groups?


### Creating a Subset

**The filter() function allows us to select observations that meet specific conditions.**

```{r}
bp_data <- data |>
  filter(group %in% c("Placebo", "High Dose"))

```


# Part 2: Summary Statistics and Visualisation

We have created `bp_data`, which contains only the High Dose and Placebo groups.

In the next activity, we will use this dataset to calculate summary statistics and create a graph.

## (1) Calculate Summary Statistics for Each Group

```{r}
group_summary <- bp_data |>
  group_by(group) |>
  summarise(
    n = n(),
    Mean_SBP = mean(sbp_post),
    SD_SBP = sd(sbp_post)
  )

group_summary
```

- The `group_by()` function tells R to split the dataset into groups based on the variable `group`.

- The `summarise()` function calculates summary statistics for each group.

  - Here we calculate:
    - `n()` = number of patients in each treatment group
    - `mean()` = average post-treatment blood pressure
    - `sd()` = standard deviation, which measures how spread out the blood pressure values are

**Questions**

- Which group has the lower mean blood pressure?

- How many patients are in each group?

- Which group shows more variability in blood pressure values?



## (2) Plotting group means

```{r}
ggplot(group_summary,
       aes(x = ,
           y =  )) +
  geom_col()
```



### Interpreting the Plot

The mean post-treatment blood pressure is:

-      for the High Dose group

-       for the Placebo group




