---
title: "Do It Yourself"
subtitle: "Session 1 Practice"
format:
  html:
    code-annotations: hover
    code-fold: false
    toc: true
    toc-depth: 2
    css: styles.css

execute:
  echo: true
  warning: false
  message: false

editor: visual
---

# Do It Yourself

Apply the skills you learned in Session 1 to answer two new research questions.

For each activity you should:

- identify the variables;
- decide whether a subset is needed;
- calculate summary statistics;
- create an appropriate graph; and
- briefly interpret your findings.

# Activity 1 (Clinical trial)

## Importing the data set

```{r}
library(tidyverse)

data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/main/datasets/sbp_demo.csv",
  show_col_types = FALSE
)

```

## Research question

> Do male and female patients differ in their post-treatment systolic blood pressure?

### Variables

- `sex` (categorical)

- `sbp_post` (numerical)

### Your task

1.  Identify the independent and dependent variables.

2.  Count the number of males and females.

3.  Calculate the mean post-treatment systolic blood pressure for each sex.

4.  Create a bar chart showing the group means.

5.  Write one or two sentences describing your findings.

# Activity 2 : Exploring a Tuberculosis Trial Dataset

## Research question

> Do patients receiving Streptomycin have higher recovery scores than patients receiving standard care?

### 🧬 Dataset Overview: `strep_tb`

**Source**: Reconstructed data from the 1948 BMJ paper\
**Study type**: Randomised, placebo-controlled trial\
**Participants**: 107 patients with tuberculosis\
**Treatment arms**:\
- `Streptomycin` (2g daily)\
- `Control` (no streptomycin)

### Variable Key

| Variable | Description |
|------------------------------------|------------------------------------|
| `patient_id` | Unique patient identifier |
| `arm` | Treatment group: Streptomycin or Control |
| `dose_strep_g` | Dose of Streptomycin in grams |
| `gender` | M / F |
| `baseline_condition` | 1 = Good, 2 = Fair, 3 = Poor |
| `baseline_temp` | Categorised body temperature |
| `baseline_esr` | ESR (inflammation marker) category |
| `baseline_cavitation` | Cavitation on X-ray: yes/no |
| `strep_resistance` | Streptomycin resistance at 6 months |
| `radiologic_6m` | X-ray outcome description |
| `rad_num` | Numeric X-ray outcome (1 = death → 6 = great improvement) |
| `improved` | TRUE if rad_num ≥ 5, otherwise FALSE |
| `recovery_score` | Continuous recovery score (0–100). Higher scores indicate better recovery. |

## Step 1: Import the dataset

```{r}

strep_data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/refs/heads/main/datasets/strep_tb_teaching.csv",
  show_col_types = FALSE
)

```

### Your task 

1.  Identify the independent variable and the dependent variable.

2.  Use count() to find out how many patients received Streptomycin and how many were in the Control group.

3.  Create a new dataset containing only the variables needed to answer the research question.

4.  Calculate the number (and, if you wish, the percentage) of patients who improved in each treatment group.

5.  Create an appropriate graph to compare the treatment groups.

6.  Write one or two sentences describing your findings.
