---
title: "Investigating a Clinical Trial Dataset"
author: " "
date: "2026"

---

# Importing the data set

```{r}
library(tidyverse)

data <- read_csv(
  "https://raw.githubusercontent.com/col-stem-uoe/Summer26/main/datasets/sbp_demo.csv",
  show_col_types = FALSE
)

```

# **Research Question A**

> Did patients in the High Dose group have lower blood pressure after treatment than before treatment?

Variables:

- `sbp_pre`

- `sbp_post`

```{r}
high_dose <- data |>
  filter(group %in% c("High Dose"))
```

or you could use `filter(group == "High Dose")` = Filter the data to keep rows where the group is High Dose.

```{r}
high_dose <- data |>
  filter(group == "High Dose")
```

```{r}
high_dose_summary <- high_dose |>
  summarise(
    n = n(),
    Mean_Pre = mean(sbp_pre),
    Mean_Post = mean(sbp_post),
    SD_Pre = sd(sbp_pre),
    SD_Post = sd(sbp_post)
  )

high_dose_summary
```

### Reshaping the Data for Plotting

The summary table produced by `summarise()` looks like this:

| Mean_Pre | Mean_Post |
|----------|-----------|
| 138.4    | 125.1     |

This format is called **wide format** because the values are stored across multiple columns.

However, `ggplot()` often works more easily when the data are stored in **long format**, where the values are arranged in rows instead.

The `pivot_longer()` function converts the table into the following format:

| Time      | Mean_SBP |
|-----------|----------|
| Mean_Pre  | 138.4    |
| Mean_Post | 125.1    |

```{r}
pivot_longer(
  cols = everything(),
  names_to = "Time",
  values_to = "Mean_SBP"
)
```

#### What does each argument do?

```         
cols = everything()
```

Select all columns in the dataset. In this case, these are:

- `Mean_Pre`

- `Mean_Post`

```         
names_to = "Time"
```

Take the column names (`Mean_Pre` and `Mean_Post`) and store them in a new variable called `Time`.

```         
values_to = "Mean_SBP"
```

Take the values stored inside those columns (`138.4` and `125.1`) and store them in a new variable called `Mean_SBP`.

```{r}
high_dose_plot <- high_dose |>
  summarise(
    Before = mean(sbp_pre),
    After = mean(sbp_post)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Time",
    values_to = "Mean_SBP"
  )
```

```{r}
high_dose_plot <- high_dose_plot |>
  mutate(
    Time = factor(
      Time,
      levels = c("Before", "After")
    )
  )
```

```{r}
ggplot(high_dose_plot,
       aes(x = Time,
           y = Mean_SBP)) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(label = round(Mean_SBP, 1)),
    vjust = -0.5
  ) +
  labs(
    x = "Time point",
    y = "Mean systolic blood pressure (mmHg)"
  )
```

# **Research Question B**

**Do male and female patients differ in their post-treatment systolic blood pressure?**

Variables:

- `sex` (categorical)

- `sbp_post` (numerical)

YOU should:

- Identify the variables.

- Use `count(data, sex)`.

- Create a subset if needed.

- Calculate group means.

- Create a bar plot.

- Interpret the results.
