---
title: "Try It Yourself: Correlation"
format:
  html:
    code-tools: true
output: 
  learnr::tutorial:
    progressive: false  #this creates the sidebar (creates it= true, turns off sidebar= flase)
    allow-skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# Install and load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, DT, medicaldata, learnr, kableExtra, ggplot2, janitor)

library(learnr)
library(dplyr)
library(ggplot2)
library(medicaldata)
library(DT)
library(janitor)
data("smartpill")
```

## 🔍 Your Task

> Test whether **gastric emptying time** is associated with **total gut transit time** in healthy volunteers using the `smartpill`dataset.

You will explore the relationship between **`gastric_emptying_time`** and **`total_gut_transit_time`** to determine whether people who take longer to empty their stomach also tend to take longer to fully digest and pass food through the entire gastrointestinal tract.

+:------------------------------------------------------------------------------------------------------------------------------+
| \### 🧬 Dataset Overview: `smartpill`                                                                                         |
+-------------------------------------------------------------------------------------------------------------------------------+
| **Source**: Data collected using the SmartPill wireless motility capsule\                                                     |
| **Study type**: Prospective cohort study\                                                                                     |
| **Participants**: 109 adults (87 healthy, 22 with gastroparesis)\                                                             |
| **Device**: SmartPill capsule (measures pH, temperature, pressure)\                                                           |
| **Primary measures**: Times (in hours) for transit through sections of the GI tract                                           |
+-------------------------------------------------------------------------------------------------------------------------------+
| For this exercise, we will focus on **healthy participants only** (N = 87) to reduce missingness and simplify interpretation. |
+-------------------------------------------------------------------------------------------------------------------------------+

### 📋 Variable Key

+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| Position | Variable                 | Variable Label                  | Units                   | Codes / Notes                                                                               |
+==========+==========================+=================================+=========================+=============================================================================================+
| 1        | `Group`                  | Study group                     | —                       | `0` = Critically Ill Trauma Patient<br>`1` = Healthy Volunteer                              |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 2        | `Gender`                 | Gender                          | —                       | `0` = Female<br>`1` = Male                                                                  |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 3        | `Race`                   | Race                            | —                       | `1` = White<br>`2` = Black<br>`3` = Asian/Pacific Islander<br>`4` = Hispanic<br>`5` = Other |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 4        | `Height`                 | Height                          | centimeters             |                                                                                             |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 5        | `Weight`                 | Weight                          | kilograms               |                                                                                             |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 6        | `Age`                    | Age                             | years                   |                                                                                             |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 7        | `GE.Time`                | Gastric Emptying Time           | hours                   | From ingestion to gastric emptying                                                          |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 8        | `SB.Time`                | Small Bowel Transit Time        | hours                   | From gastric emptying to ileocecal junction                                                 |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 9        | `C.Time`                 | Colonic Transit Time            | hours                   | From ileocecal junction to body exit                                                        |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 10       | `WG.Time`                | Whole Gut Transit Time          | hours                   | From ingestion to body exit                                                                 |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 11       | `S.Contractions`         | Stomach Contractions            | contractions per minute | Peaks over 10 mmHg and under 300 mmHg                                                       |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 12       | `S.Sum.of.Amplitudes`    | Stomach Sum of Amplitudes       | —                       |                                                                                             |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 13       | `S.Mean.Peak.Amplitude`  | Stomach Mean Peak Amplitude     | mm Hg                   | Sum of amplitudes ÷ number of contractions                                                  |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 14       | `S.Mean.pH`              | Stomach Mean pH                 | —                       | Average over recording time                                                                 |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 15       | `SB.Contractions`        | Small Bowel Contractions        | contractions per minute | Peaks over 10 mmHg and under 300 mmHg                                                       |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 16       | `SB.Sum.of.Amplitudes`   | Small Bowel Sum of Amplitudes   | —                       |                                                                                             |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 17       | `SB.Mean.Peak.Amplitude` | Small Bowel Mean Peak Amplitude | mm Hg                   | Sum of amplitudes ÷ number of contractions                                                  |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 18       | `SB.Mean.pH`             | Small Bowel Mean pH             | —                       | Average over recording time                                                                 |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 19       | `C.Contractions`         | Colon Contractions              | contractions per minute | Peaks over 10 mmHg and under 300 mmHg                                                       |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 20       | `C.Sum.of.Amplitudes`    | Colon Sum of Amplitudes         | —                       |                                                                                             |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 21       | `C.Mean.Peak.Amplitude`  | Colon Mean Peak Amplitude       | mm Hg                   | Sum of amplitudes ÷ number of contractions                                                  |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+
| 22       | `C.Mean.pH`              | Colon Mean pH                   | —                       | Average over recording time                                                                 |
+----------+--------------------------+---------------------------------+-------------------------+---------------------------------------------------------------------------------------------+

### Step 1: Preview the Dataset

#### Raw Dataset

```{r raw_dataset, results='asis', exercise=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(DT)

# Show structure
glimpse(smartpill)

```

#### Formatted Dataset

```{r interactive_dataset, results='asis', exercise=TRUE, warning=FALSE, message=FALSE}
DT::datatable(
  smartpill,
  options = list(scrollX = TRUE),
  class = 'stripe nowrap',
  caption = "Preview of the smartpill Dataset"
)

```

### Step 2: Summarise by Group

#### Healthy Patients

```{r summary_by_group_healthy, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}
# Your turn! Fill in the missing filters

# Filter for healthy volunteers (Group = 1)
healthy_data <- smartpill |>
  filter(Group == 1)

# Preview the dataset
DT::datatable(
  healthy_data,
  options = list(scrollX = TRUE),
  class = 'stripe nowrap',
  caption = "Preview of the Healthy Volunteers Dataset"
)

```

#### Critically Ill Patients

```{r summary_by_group_gp, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}

# Filter for trauma patients (Group = 0)
critically_ill_data <- smartpill |>
  filter(Group == 0)

# Preview the dataset
DT::datatable(
  critically_ill_data,
  options = list(scrollX = TRUE),
  class = 'stripe nowrap',
  caption = "Preview of the Trauma Patients Dataset"
)

```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br> 1. Replace the first blank with <code>Group</code> — this is the variable that indicates patient type.<br> 2. Use <code>1</code> for Healthy volunteers and <code>0</code> for Critically Ill Trauma patients.<br> 3. Use <code>filter()</code> to isolate just the participants you want.<br> 4. The filtered datasets (e.g. <code>healthy_data</code>, <code>gp_data</code>) will be used in later steps for plotting and correlation analysis.
:::

### Step 3: Visualise the Relationship

We’re going to explore whether people who take longer to empty their stomachs also take longer to fully pass food through the digestive system.

For this step, you’ll create a scatterplot to visualise the relationship between:

GE.Time: Gastric Emptying Time — how long it takes for food to leave the stomach WG.Time: Whole Gut Transit Time — how long it takes from ingestion to excretion We'll look at this relationship separately for each group:

> Healthy volunteers (Group = 1)
>
> Critically ill trauma patients (Group = 0)

#### 3a. Healthy Volunteers

🧪 **Research Question:**\
*Among healthy participants, is there a relationship between gastric emptying time and total gut transit time?*

```{r scatterplot, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}

# Filter for healthy volunteers (Group = 1)
healthy_data <- smartpill |>
  filter(Group == 1)

# Your turn! Plot the relationship in healthy volunteers

ggplot(healthy_data, aes(x = ____, y = ____)) +      # <1>
  geom_point(colour = "#C5004A", size = 2.5) +        # <2>
  geom_smooth(method = "lm", se = FALSE, colour = "black") +  # <3>
  labs(title = "Gastric Emptying vs Total Gut Transit Time",
       x = "Gastric Emptying Time (hours)",
       y = "Total Gut Transit Time (hours)") +
  theme_minimal(base_size = 14)                      # <4>

```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br>

Replace the blanks with the appropriate numeric variables: • <code>x = gastric_emptying_time</code> • <code>y = total_gut_transit_time</code><br> Use <code>geom_point()</code> to plot each patient as a dot.<br> Add <code>geom_smooth(method = "lm")</code> for a linear trend line (Pearson correlation).<br> Use clear axis labels and a minimal theme for readability.
:::

```{r scatterplot-solution}
# Filter for healthy volunteers (Group = 1)
healthy_data <- smartpill |>
  filter(Group == 1)

# Your turn! Plot the relationship in healthy volunteers
ggplot(healthy_data, aes(x = `GE.Time`, y = `WG.Time`)) +
  geom_point(colour = "#C5004A", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(title = "Gastric Emptying vs Total Gut Transit Time",
       x = "Gastric Emptying Time (hours)",
       y = "Total Gut Transit Time (hours)") +
  theme_minimal(base_size = 14)

```

#### 3b. Critically Ill Patients

```{r scatterplot_ill, exercise=TRUE, results='asis', warning=FALSE, message=FALSE }

# Filter for critically ill patients (Group = 0)
critically_ill_data <- smartpill |> 
  filter(Group == 0)

# Plot the relationship
ggplot(___, aes(x = `___`, y = `___`)) +
  g geom_point(colour = "#2B5D8C", size = 2.5) +  # Dark blue points
  geom_smooth(method = "lm", se = FALSE, colour = "#1B9E77") +  # Dark green regression line
  labs(title = "Gastric Emptying vs Total Gut Transit Time (Critically Ill)",
       x = "Gastric Emptying Time (hours)",
       y = "Total Gut Transit Time (hours)") +
  theme_minimal(base_size = 14)

```

```{r scatterplot_ill-solution}

# Filter for critically ill patients (Group = 0)
critically_ill_data <- smartpill |> 
  filter(Group == 0)

# Plot the relationship
ggplot(critically_ill_data, aes(x = `GE.Time`, y = `WG.Time`)) +
  geom_point(colour = "#2B5D8C", size = 2.5) +  # Dark blue points
  geom_smooth(method = "lm", se = FALSE, colour = "#1B9E77") +  # Dark green regression line
  labs(title = "Gastric Emptying vs Total Gut Transit Time (Critically Ill)",
       x = "Gastric Emptying Time (hours)",
       y = "Total Gut Transit Time (hours)") +
  theme_minimal(base_size = 14)


```

### EXTRA: Compare Groups

```{r plot_comparegroups, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}

# Convert Group to factor for clearer labels
smartpill$Group <- factor(smartpill$Group,
                          levels = c(0, 1),
                          labels = c("Critically Ill", "Healthy"))

# Plot both groups together, colour-coded
ggplot(smartpill, aes(x = `GE.Time`, y = `WG.Time`, colour = Group)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_colour_manual(values = c("Critically Ill" = "#2B5D8C",
                                 "Healthy" = "#C5004A")) +
  labs(title = "Gastric Emptying vs Total Gut Transit Time by Group",
       x = "Gastric Emptying Time (hours)",
       y = "Total Gut Transit Time (hours)",
       colour = "Group") +
  theme_minimal(base_size = 14)

```
::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br>
- Use the full <code>smartpill</code> dataset to include all participants.<br>
- Set <code>colour = Group</code> inside <code>aes()</code> to colour the dots by group.<br>
- Add <code>geom_smooth()</code> to show a trend line per group.<br>
- Use <code>scale_colour_manual()</code> to customise colours (optional).<br>
- Label axes clearly and include a legend title.
:::


<div style="border-left: 4px solid #1976D2; padding: 1em; background-color: #E3F2FD; border-radius: 4px;">
<strong>🧠 Think about it...</strong><br><br>
• Does the relationship between gastric emptying and total gut transit time look different across groups?<br>
• Which group shows a stronger correlation?<br>
• What does this suggest about digestion in healthy vs critically ill patients?
</div>


### Step 4: Run a Correlation Test
#### 4a. Healthy Volunteers

```{r run-correlation-healthy, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}
# Filter again just in case
healthy_data <- smartpill |>
  filter(Group == 1)

# Your turn! Run a Pearson correlation test
cor.test(___, ___, method = "pearson")   # <1>

```
::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br>

Use <code>healthy_data$GE.Time</code> as the first argument (X variable)<br>
Use <code>healthy_data$WG.Time</code> as the second argument (Y variable)<br>
Make sure the method is set to <code>"pearson"</code>
:::

```{r run-correlation-healthy-solution}
# Filter again just in case
healthy_data <- smartpill |>
  filter(Group == 1)

# Pearson correlation for healthy volunteers
cor.test(healthy_data$`GE.Time`, healthy_data$`WG.Time`, method = "pearson")

```

#### 4b. Critically Ill Patients
```{r run-correlation-ill, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}
# Filter for critically ill patients
critically_ill_data <- smartpill |>
  filter(Group == 0)

# Your turn! Run a Pearson correlation test
cor.test(___, ___, method = "pearson")   # <1>

```
::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br>

Use <code>critically_ill_data$GE.Time</code> and <code>critically_ill_data$WG.Time</code><br>
This will tell us if gastric emptying predicts total gut transit time among critically ill patients
:::

```{r run-correlation-ill-solution}
# Filter for critically ill patients
critically_ill_data <- smartpill |>
  filter(Group == 0)

# Pearson correlation for critically ill patients
cor.test(critically_ill_data$`GE.Time`, critically_ill_data$`WG.Time`, method = "pearson")

```

### Step 5: Interpret the Results

Let’s reflect on what the correlation results mean for each group.

#### 5a. Healthy Volunteers

Use multiple-choice and fill-in-the-blank to help students extract the key values from the test output.

```{r interpretation_question, echo=FALSE}
question("What was the correlation coefficient (r) for the healthy group?",
  answer("0.25", correct = TRUE),
  answer("0.52"),
  answer("–0.38"),
  answer("0.91"),
  allow_retry = TRUE
)

```

```{r interpretation_question2, echo=FALSE}
question("Was the correlation statistically significant (p < .05)?",
  answer("Yes ✅", correct = TRUE),
  answer("No ❌"),
  allow_retry = TRUE
)

```

Now, write your interpretation here:

```{r writeup-interpretation, exercise=TRUE}
# Write your interpretation here

# What does the correlation mean in the context of healthy volunteers?
# Think about direction, strength, significance, and what it implies about the relationship.


```

```{r writeup-interpretation-solution}
Among healthy participants, there was a **weak but statistically significant positive correlation** between gastric emptying time and total gut transit time (r = 0.25, p = .020). This suggests that those who take longer to empty their stomach also tend to have longer overall digestion times.

```


#### 5b. Critically Ill Patients

```{r interpretation_question_ill, echo=FALSE}
question("What was the correlation coefficient (r) for the critically ill group?",
  answer("0.52", correct = TRUE),
  answer("0.25"),
  answer("–0.38"),
  answer("0.91"),
  allow_retry = TRUE
)

```


```{r interpretation_question_ill2, echo=FALSE}
question("Was the correlation statistically significant (p < .05)?",
  answer("No ❌", correct = TRUE),
  answer("Yes ✅"),
  allow_retry = TRUE
)

```

Now, write your interpretation below:

```{r writeup-interpretation-ill, exercise=TRUE}
# Write your interpretation here

# What does the correlation mean in the context of critically ill patients?
# Think about direction, strength, significance, and what it implies about the relationship.

```

```{r writeup-interpretation-ill-solution}
Among critically ill patients, there was a **moderate positive correlation** between gastric emptying time and total gut transit time (r = 0.52), but this result was **not statistically significant** (p = .231). This means we cannot conclude that the two variables are reliably related in this group — the observed association could be due to chance, especially given the small sample size.

```


::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint: How to interpret a correlation</strong><br><br>

- **Direction**:  
  A **positive** correlation means that both variables increase together.  
  A **negative** correlation means that as one increases, the other decreases.

- **Strength**:  
  Use the correlation coefficient **r** to describe strength:  
  • 0.00–0.19 = Very weak  
  • 0.20–0.39 = Weak  
  • 0.40–0.59 = Moderate  
  • 0.60–0.79 = Strong  
  • 0.80–1.00 = Very strong  

- **Significance**:  
  If the **p-value is below 0.05**, the result is **statistically significant** — meaning the observed relationship is unlikely to be due to chance.

✅ Use all three elements — direction, strength, and significance — to interpret the result clearly.
:::

### ✍️ Step 6: Write-up Across Groups
Now it’s your turn to write a short results paragraph using the correlation results you interpreted above.

Use the template below and replace the blanks with your results:

```{r  writeup-interpretation-general, exercise=TRUE}


# Complete the paragraph below by replacing the blanks with your findings from Step 5.

# Remember to include: the direction, strength, significance, and a comparison between groups.

This study examined whether ________ was associated with ________ in both ________ and ________ participants.

Among healthy participants, there was a ________ (strength) and ________ (significance) correlation between the two variables (r = ____, p = ____), suggesting that ________.

Among critically ill patients, the association was ________ (strength) and ________ (significance) (r = ____, p = ____), indicating that ________.

Together, these results suggest that while the association exists in both groups, it is ________ (comparative statement), possibly due to ________.

```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br>
Include the following elements:<br><br>

Who the results are about (Healthy or Critically Ill patients)
Direction and strength of the correlation (positive/negative; weak/moderate/etc.)
The correlation coefficient (r) and the p-value
A sentence about what the result means in context
:::

```{r writeup-interpretation-general-solution}

This study examined whether gastric emptying time was associated with total gut transit time in both healthy and critically ill participants.

Among healthy participants, there was a weak but statistically significant correlation between the two variables (r = 0.25, p = .020), suggesting that those who take longer to empty their stomach also tend to have longer digestive times.

Among critically ill patients, the association was moderate and statistically significant (r = 0.46, p = .029), indicating that delayed gastric emptying is more strongly linked to total transit time in this group.

Together, these results suggest that while the association exists in both groups, it is stronger in critically ill patients, possibly due to impaired gastrointestinal motility in critical illness.

```

