---
title: "Try It Yourself: Test a Hypothesis t-test"
format:
  html:
    code-tools: true
output: 
  learnr::tutorial:
    progressive: false  #this creates the sidebar (creates it= true, turns off sidebar= flase)
    allow-skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# Install and load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, DT, medicaldata, learnr, kableExtra, ggplot2)

library(learnr)
library(gradethis)
library(dplyr)
library(ggplot2)
library(medicaldata)
data("strep_tb")
```

## 🔍 Your Task

> Compare **radiologic scores** between **treatment** and **control** groups using the `strep_tb` dataset.

------------------------------------------------------------------------

### 🧬 Dataset Overview: `strep_tb`

**Source**: Reconstructed data from the 1948 BMJ paper\
**Study type**: Randomised, placebo-controlled trial\
**Participants**: 107 patients with tuberculosis\
**Treatment arms**:\
- `Streptomycin` (2g daily)\
- `Control` (no streptomycin)

**Outcome of interest**:\
- `improved`: Logical variable (TRUE/FALSE) derived from radiologic score\
- TRUE = radiologic score of 5 or 6 (moderate/considerable improvement)

------------------------------------------------------------------------

### 📋 Variable Key

| Variable | Description |
|------------------------------------|------------------------------------|
| `patient_id` | Unique patient identifier |
| `arm` | Treatment group: Streptomycin or Control |
| `dose_strep_g` | Dose of Streptomycin in grams |
| `gender` | M / F |
| `baseline_condition` | 1 = Good, 2 = Fair, 3 = Poor |
| `baseline_temp` | Categorised body temperature |
| `baseline_esr` | ESR (inflammation marker) category |
| `baseline_cavitation` | Cavitation on X-ray: yes/no |
| `strep_resistance` | Streptomycin resistance at 6 months |
| `radiologic_6m` | X-ray outcome description |
| `rad_num` | Numeric X-ray outcome (1 = death → 6 = great improvement) |
| `improved` | TRUE if rad_num ≥ 5, otherwise FALSE |

------------------------------------------------------------------------

### Step 1: Preview the Dataset

#### Raw Dataset

```{r raw_dataset, results='asis', exercise=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(DT)
library(medicaldata)

# Show structure
glimpse(strep_tb)

```

#### Formatted Dataset

```{r interactive_dataset, results='asis', exercise=TRUE, warning=FALSE, message=FALSE}
DT::datatable(
  strep_tb,
  options = list(scrollX = TRUE),
  class = 'stripe nowrap',
  caption = "Preview of the strep_tb Dataset"
)

```

### Step 2: Summarise by Group

```{r summary_by_group, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}

# Your turn! Fill in the missing variable names
library(dplyr)
library(kableExtra)

summary_result <- strep_tb |> 
  group_by(_____) |>                        # <1>
  summarise(
    mean_score = mean(_____, na.rm = TRUE), # <2>
    sd_score = sd(_____, na.rm = TRUE),     # <2>
    se_score = sd_score / sqrt(n())
  )


kable(summary_result,
      caption = "📊 Summary of Radiologic Scores by Group") |>
  kable_styling(full_width = FALSE, position = "center", font_size = 16)

```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br> 1. Replace the first blank with <code>arm</code> – this is your grouping variable (x-axis).<br> 2. Replace the second and third blanks with <code>rad_num</code> – this is your outcome variable (y-axis).<br> 3. Use <code>na.rm = TRUE</code> inside both <code>mean()</code> and <code>sd()</code> to exclude missing values.<br> 4. The <code>se_score</code> (standard error) is calculated by dividing the <code>sd_score</code> by the square root of the sample size in each group:\
<code>se_score = sd_score / sqrt(n())</code>.
:::

```{r summary_by_group-solution}
library(dplyr)
library(kableExtra)

summary_result <- strep_tb |> 
  group_by(arm) |> 
  summarise(
    mean_score = mean(rad_num, na.rm = TRUE),
    sd_score = sd(rad_num, na.rm = TRUE),
    se_score = sd_score / sqrt(n())
  )

kable(summary_result,
      caption = "🧮 Summary of Radiologic Scores by Group") |> 
  kable_styling(full_width = FALSE, position = "center", font_size = 16)
```

### Step 3: Visualise the Comparison

```{r plot_group_means, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}

# 1. Re-create summary_result inside this chunk




#  2. Your turn! Replace the blanks with the correct variable names
# Use your summary table from Step 2

ggplot(_____, aes(x = _____, y = _____)) +
  geom_col(fill = "#C5004A", colour = "black", width = 0.6) +
  geom_errorbar(aes(ymin = _____ - _____, ymax = _____ + _____), width = 0.2) +
  geom_text(aes(label = round(_____, 1)), 
            vjust = -1.2, size = 5, fontface = "bold", colour = "black") +
  labs(title = "Radiologic Score by Group",
       x = "Group", y = "Mean Radiology Score (1–6)") +
   ylim(0,6)+
  theme_minimal(base_size = 14)
```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br> - Use <code>summary_result</code> from Step 2<br> as the first argument of ggplot - <code>arm</code> is the grouping variable (x-axis)<br> - <code>mean_score</code> is your y-axis<br> - <code>se_score</code> gives the error bar height<br> - Add <code>geom_text()</code> to print the mean score on top of each bar
:::

```{r plot_group_means-solution}

# 1. Re-create summary_result inside this chunk
summary_result <- strep_tb |>
  group_by(arm) |>
  summarise(
    mean_score = mean(rad_num, na.rm = TRUE),
    sd_score = sd(rad_num, na.rm = TRUE),
    se_score = sd_score / sqrt(n())
  )

# 2. Then run the plot
ggplot(summary_result, aes(x = arm, y = mean_score)) +
  geom_col(fill = "#C5004A", colour = "black", width = 0.6) +
  geom_errorbar(aes(ymin = mean_score - se_score, ymax = mean_score + se_score), width = 0.2) +
  geom_text(aes(label = round(mean_score, 1)),
            vjust = -1.2, size = 5, fontface = "bold", colour = "black") +
  labs(title = "Radiologic Score by Group",
       x = "Group", y = "Mean Radiology Score (1–6)") +
   ylim(0,6)+
  theme_minimal(base_size = 14)

```

### Step 4: Run a t-test

```{r run_ttest, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}
t.test(____ ~ ____, data = strep_tb)

```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong>\
Use <code>rad_num</code> as the dependent variable (DV), and <code>arm</code> as the independent variable (IV).\
The formula should be written as: <code>DV \~ IV</code>.
:::

```{r run_ttest-solution}
t.test(rad_num ~ arm, data = strep_tb)
```

### ✍️ Step 5: Interpretation

Use the output from your t-test in Step 4 to decide whether the difference is **statistically significant**.\
Focus on the **p-value**, which tells you whether the group difference is likely due to chance.

```{r interpretation_question, echo=FALSE}
question("Was the difference significant?",
  answer("Yes – the p-value was below 0.05", correct=TRUE),
  answer("No – the means were equal"),
  answer("No – sample size was too small"),
  allow_retry = TRUE
)
```

### ✍️ Write-Up Practice

Now it’s your turn to **write a short results paragraph** using your own t-test output.

Use the template below and **replace the blanks** with your results:

``` writeup-hint
An independent samples t-test found that patients in the _____ group (M = ___, SD = ___) had significantly _____ post-treatment radiologic scores than those in the _____ group (M = ___, SD = ___), t(___) = ___, p = ___.
```

```{r writeup-draft, exercise=TRUE}
# Write your paragraph here using your own results
```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br> - Mean and SD for each group<br> - The t-value and degrees of freedom<br> - The p-value (rounded to 3 decimal places)<br> - A conclusion on whether the difference was significant
:::

```{r writeup-draft-solution}
An independent samples t-test found that patients in the Streptomycin group (M = 4.67, SD = 1.73) had significantly higher post-treatment radiologic scores than those in the Control group (M = 3.13, SD = 1.73), t(104.7) = 4.60, p < .001. This suggests that the treatment group had better radiologic outcomes than the control group, indicating that Streptomycin did improve lung damage on X-rays compared to standard care alone.

```
