---
title: "Try It Yourself: Chi-square Test"
format:
  html:
    code-tools: true
output: 
  learnr::tutorial:
    progressive: false  #this creates the sidebar (creates it= true, turns off sidebar= flase)
    allow-skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# Install and load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, DT, medicaldata, learnr, kableExtra, ggplot2, janitor)

library(learnr)
library(dplyr)
library(ggplot2)
library(medicaldata)
library(DT)
library(janitor)
data("smartpill")
```

## 🔍 Your Task

> Test whether **treatment group** is associated with **clinical improvement** in patients with tuberculosis using the `strep_tb` dataset.

You will analyse whether receiving **Streptomycin** is linked to a higher chance of **clinical improvement**, as measured by radiologic outcomes. Specifically, you’ll test the association between the **treatment group** (`arm`) and a new binary **outcome** variable (`improved`), which indicates whether patients showed moderate or considerable improvement on their 6-month X-ray.

To do this, you’ll:

- Create a contingency table showing counts of improved/not improved patients by treatment group.
- Visualise the data using a **stacked bar chart**.
- Run a **chi-square test of independence**.
- Write up your results with the appropriate statistical summary.



------------------------------------------------------------------------

### 🧬 Dataset Overview: `strep_tb`

**Source**: Reconstructed data from the 1948 BMJ paper\
**Study type**: Randomised, placebo-controlled trial\
**Participants**: 107 patients with tuberculosis\
**Treatment arms**:\
- `Streptomycin` (2g daily)\
- `Control` (no streptomycin)

**Outcome of interest**:\
- `improved`: Logical variable (TRUE/FALSE) derived from radiologic score\
- TRUE = radiologic score of 5 or 6 (moderate/considerable improvement)

------------------------------------------------------------------------

### 📋 Variable Key

| Variable | Description |
|------------------------------------|------------------------------------|
| `patient_id` | Unique patient identifier |
| `arm` | Treatment group: Streptomycin or Control |
| `dose_strep_g` | Dose of Streptomycin in grams |
| `gender` | M / F |
| `baseline_condition` | 1 = Good, 2 = Fair, 3 = Poor |
| `baseline_temp` | Categorised body temperature |
| `baseline_esr` | ESR (inflammation marker) category |
| `baseline_cavitation` | Cavitation on X-ray: yes/no |
| `strep_resistance` | Streptomycin resistance at 6 months |
| `radiologic_6m` | X-ray outcome description |
| `rad_num` | Numeric X-ray outcome (1 = death → 6 = great improvement) |
| `improved` | TRUE if rad_num ≥ 5, otherwise FALSE |

------------------------------------------------------------------------

### Step 1: Preview the Dataset

#### Raw Dataset

```{r raw_dataset, results='asis', exercise=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(DT)

# Show structure
glimpse(strep_tb)

```

#### Formatted Dataset

```{r interactive_dataset, results='asis', exercise=TRUE, warning=FALSE, message=FALSE}
DT::datatable(
  strep_tb,
  options = list(scrollX = TRUE),
  class = 'stripe nowrap',
  caption = "Preview of the strep_tb Dataset"
)

```

### Step 2: Summarise by Group

```{r summary_by_group, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}


# Your turn! Fill in the missing variable names
library(dplyr)
library(kableExtra)

# Step 0: Set variable order and create outcome variable based on improvement
strep_tb$outcome <- ifelse(strep_tb$improved == TRUE, "Yes", "No")
strep_tb$outcome <- factor(strep_tb$outcome, levels = c("No", "Yes"))

# Step 1: Count and calculate %
label_data <- strep_tb |> 
  count(_____, _____) |>          # <1>
  group_by(_____) |>              # <2>
  mutate(
    percent = n / sum(n),         # <3>
    label = paste0(round(percent * 100), "%")  # <4>
  )

label_data

```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br> 1. Replace both blanks with the correct variables: <code>treatment group</code> and <code>outcome</code>.<br> 2. Group by the treatment variable so the percentages add to 100% within each group.<br> 3. Compute the proportion of each outcome level (<code>Yes</code>/<code>No</code>) per group.<br> 4. Create a text label like <code>"80%"</code> to use in the next step (plotting).
:::

```{r summary_by_group-solution}

# Solution: summarise categorical outcome by group

# Step 0: Set variable order and create outcome variable based on improvement
strep_tb$outcome <- ifelse(strep_tb$improved == TRUE, "Yes", "No")
strep_tb$outcome <- factor(strep_tb$outcome, levels = c("No", "Yes"))

# Step 1: Count and calculate %
label_data <- strep_tb |>
  count(arm, outcome) |>
  group_by(arm) |>
  mutate(
    percent = n / sum(n),
    label = paste0(round(percent * 100), "%")
  )

label_data

```

### Step 3: Visualise the Comparison (Stacked Bar Chart)

::: {style="border-left: 4px solid #1E88E5; padding: 1em; background-color: #E3F2FD; border-radius: 4px;"}
<strong>📌 Important:</strong><br><br>

<code>strep_tb</code> is your **main dataset** for <code>ggplot()</code>:\
Use it to build the stacked bars with <code>geom_bar(position = "fill")</code>.\
This will automatically count values and scale each bar to 100%.

<code>label_data</code> is used **only inside** <code>geom_text()</code>:\
It contains your pre-computed percentages and formatted labels (e.g. <code>"80%"</code>) from Step 2.

Summary:

<ul style="margin-top: 0.5em; margin-bottom: 0;">

<li><code>strep_tb</code> = raw data → used for bar height + counts</li>

<li><code>label_data</code> = summarised data → used for label positioning</li>

</ul>
:::

```{r plot_stack_comparison, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}

# 1. Re-create summary_result inside this chunk



# 2. Your turn! Use your table from Step 2
library(ggplot2)

ggplot(_____, aes(x = _____, fill = _____)) +                 # <1>
  geom_bar(position = "fill", colour = "black") +             # <2>
  geom_text(data = _____,                                     # <3>
            aes(x = _____, y = _____, label = _____, fill = _____),
            position = position_stack(vjust = 0.5),
            colour = "white", size = 4.5, fontface = "bold") +
  scale_y_continuous(labels = scales::percent) +              # <4>
  labs(title = "Proportion of Patients Who Improved by Group",
       y = "Percentage", x = "Treatment Group",
       fill = "Improved?") +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_line(colour = "grey85"),
    panel.grid.minor = element_blank()
  )

```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br> • Use <code>strep_tb</code> as your main dataset for <code>ggplot()</code><br> • Set <code>x = arm</code> and <code>fill = outcome</code> inside <code>aes()</code><br> • Use the <code>label_data</code> table from Step 2 for <code>geom_text()</code><br> • In <code>geom_text()</code>, map <code>y = percent</code> and <code>label = label</code><br> • Add <code>position = position_stack(vjust = 0.5)</code> to centre the labels<br> • Format y-axis using <code>scale_y_continuous(labels = scales::percent)</code>
:::

```{r plot_stack_comparison-solution}

# 1. Re-create summary_result inside this chunk
strep_tb$outcome <- ifelse(strep_tb$improved == TRUE, "Yes", "No")
strep_tb$outcome <- factor(strep_tb$outcome, levels = c("No", "Yes"))

label_data <- strep_tb |> 
  count(arm, outcome) |> 
  group_by(arm) |> 
  mutate(
    percent = n / sum(n),
    label = paste0(round(percent * 100), "%")
  )

label_data

# 2. Your turn! Use your table from Step 2
ggplot(strep_tb, aes(x = arm, fill = outcome)) +
  geom_bar(position = "fill", colour = "black") +
  geom_text(data = label_data,
            aes(x = arm, y = percent, label = label, fill = outcome),
            position = position_stack(vjust = 0.5),
            colour = "white", size = 4.5, fontface = "bold") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Patients Who Improved by Group",
       y = "Percentage", x = "Treatment Group",
       fill = "Improved?") +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_line(colour = "grey85"),
    panel.grid.minor = element_blank()
  )

```

### Step 4: Run a chi-square test

```{r run_chisquare, exercise=TRUE, results='asis', warning=FALSE, message=FALSE}
# We need to create the column "outcome again"

strep_tb$outcome <- ifelse(strep_tb$improved == TRUE, "Yes", "No")
strep_tb$outcome <- factor(strep_tb$outcome, levels = c("No", "Yes"))

# Step 1: Create contingency table
table_outcome <- table(strep_tb$arm, strep_tb$outcome)

# Step 2: Run chi-square test of independence
# Your turn! Run a chi-square test of independence
chi_test <- chisq.test(___)

# Step 3: View results
chi_test
```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br>

1.  <code>ifelse(…)</code> is used to create a new <code>outcome</code> column based on whether each patient improved – this turns <code>TRUE/FALSE</code> into <code>"Yes"/"No"</code>.<br>
2.  <code>factor(…)</code> makes sure the outcome variable has the correct order: <code>No</code> first, then <code>Yes</code>.<br>
3.  <code>table(strep_tb\$arm, strep_tb\$outcome)</code> creates a contingency table showing how many patients improved in each treatment group.<br>
4.  <code>chisq.test(…)</code> runs a chi-square test on this table to check if there’s a relationship between treatment and improvement.<br><br>

✅ The result will tell you if there’s a statistically significant association between treatment and improvement.
:::

```{r run_chisquare-solution}
# We need to create the column "outcome again"

strep_tb$outcome <- ifelse(strep_tb$improved == TRUE, "Yes", "No")
strep_tb$outcome <- factor(strep_tb$outcome, levels = c("No", "Yes"))

# Step 1: Create contingency table
table_outcome <- table(strep_tb$arm, strep_tb$outcome)

# Step 2: Run chi-square test of independence
chi_test <- chisq.test(table_outcome)

# Step 3: View results
chi_test
```

✅ The test checks whether improvement is **related** to the treatment group, based on a **contingency table of counts**.

You’ll see three important pieces of output:

| Term | Meaning |
|------------------------------------|------------------------------------|
| `X-squared` | The test statistic — how different the observed counts are from expected |
| `df` | Degrees of freedom (based on the number of rows/columns in your table) |
| `p-value` | Whether the observed association is statistically significant |

If the **p-value is below 0.05**, we reject the null hypothesis and conclude:

> “There is a significant association between treatment group and clinical outcome.”

This means that **treatment may influence whether patients improved**.\

### ✍️ Step 5: Interpretation

```{r interpretation_question, echo=FALSE}
question("Was the association between treatment and outcome significant?",
  answer("Yes – the p-value was below 0.05", correct = TRUE),
  answer("No – there was no association", correct = FALSE),
  answer("No – the test was not appropriate for categorical data", correct = FALSE),
  allow_retry = TRUE
)
```

> 💡 **Note:**\
> A post hoc test is not necessary here because both variables — treatment group (`arm`) and outcome (`improved`) — only have two levels each.\
> The chi-square test already compares the only available pair: Control vs Streptomycin.\
> Post hoc tests are useful when you have three or more groups and need to identify which pairs differ.\
> In this case, the main chi-square result gives you all the information you need.

### ✍️ Step 6: Write-Up Practice

Now it’s your turn to **write a short results paragraph** using your Chi-square output.

Use the template below and **replace the blanks** with your results:

``` writeup-hint

A Chi-square test of independence showed a significant association between _____ and _____, χ²(___, N = ___) = ___, p = ___. These results suggest that _____.
```

```{r writeup-draft, exercise=TRUE}
# Write your paragraph here using your own results
```

::: {style="border-left: 4px solid #D3A100; padding: 1em; background-color: #FFF8E1; border-radius: 4px;"}
<strong>💡 Hint:</strong><br><br> Include the following in your write-up:<br><br> - The two categorical variables you tested<br> - The Chi-square statistic (χ²), degrees of freedom (df), and sample size (N)<br> - The p-value (rounded to 3 decimal places)<br> - A short conclusion on whether the variables were related
:::

```{r writeup-draft-solution}
This study examined whether **treatment group** was associated with **patient outcome** in a 1948 tuberculosis trial.

As shown in Figure 1 (plot), the **Streptomycin group** had a higher proportion of patients who improved compared to the **Control group**.

A Chi-square test of independence showed a significant association between treatment group and clinical outcome, χ²(1, N = 107) = 11.48, p = .001. These findings suggest that patients receiving Streptomycin were more likely to show clinical improvement than those who received only bed rest.

```
