---
title: "Worked Example"
format:
  html:
    code-annotations: hover
    code-fold: true
    code-fold-label: "Show Code"
    toc: true
    toc-depth: 3
    css: styles.css
editor: visual
---

# 🎓 Worked Example: Analysing Real Data

This example walks through a complete analysis using real clinical data.\
We’ll guide you step by step — from research question to final interpretation — using clear code, plots, and explanations.

------------------------------------------------------------------------

## Dataset: `sbp_demo.csv`

The dataset is a fictional simulation of a clinical trial testing a blood pressure treatment.

- **Treatment groups**:
  - Placebo\
  - Low Dose Hypotensol (25mg)\
  - High Dose Hypotensol (50mg)
- **Outcome variable**:
  - `sbp_post`: Systolic Blood Pressure **after treatment** (mmHg)

------------------------------------------------------------------------

### 🎯 Research Question

> Does the **dose of treatment** affect post-treatment systolic blood pressure?

------------------------------------------------------------------------

## Step 1: Load and Prepare Data

```{r}
library(tidyverse)

# Load the dataset
sbp_demo <- read_csv("sbp_demo.csv")

# Preview
glimpse(sbp_demo)
```

## Step 2: Summary Statistics by Group

```{r}
# Summarise post-treatment SBP by group
summary_table <- sbp_demo |>
  group_by(group) |>
  summarise(
    n = n(),
    mean = round(mean(sbp_post), 1),
    median = round(median(sbp_post), 1),
    sd = round(sd(sbp_post), 1),
    min = round(min(sbp_post), 1),
    max = round(max(sbp_post), 1)
  )

summary_table

```

## Step 3: Visualise the Data

```{r}
# Boxplot of post-treatment SBP by group
ggplot(sbp_demo, aes(x = group, y = sbp_post)) +
  geom_boxplot(fill = "#C5004A", alpha = 0.6) +
  labs(title = "Post-Treatment SBP by Group",
       x = "Treatment Group", y = "SBP (mmHg)") +
  theme_minimal(base_size = 14)

```

## Step 4: Run a One-Way ANOVA

```{r}
# Run the ANOVA model
anova_result <- aov(sbp_post ~ group,data = sbp_demo)                           
anova_result
summary(anova_result)                          
```

```{r}
# Tukey HSD to compare pairs of groups
TukeyHSD(anova_result)

```

## Step 6: Interpretation

::: {.callout-note appearance="simple" icon="info" title="🧠 Summary of Results"}
There was a statistically significant difference between groups (p \< .001). The High Dose group had significantly lower SBP than both Placebo and Low Dose. This suggests the treatment is effective, especially at higher doses.
:::

## ✅ Key Takeaways

- We used ANOVA to compare **mean SBP** across 3 treatment groups.

- We followed up with **Tukey’s test** to identify where the differences lie.

- Visualisations helped us understand the spread and pattern in the data.

- Clear statistical reporting supports strong scientific communication.
